\documentclass[../../EMT-169.tex]{subfiles}
\input{../ch09/ch09.tex}

\begin{document}
\setcounter{chapter}{10}
\label{ch:chapter11}
\clearpage
	
% Glossary acronym entries %
	\newacronym{im}{IM}{intramuscular}
	\newacronym{in}{IN}{intranasal}
	\newacronym{io}{IO}{intraosseous}
	% acronym {iv} defined in ch01
	\newacronym{mdi}{MDI}{metered-dose inhaler}
	\newacronym{mi}{MI}{myocardial infarction}
	\newacronym{otc}{OTC}{over-the-counter}
	\newacronym{po}{PO}{per oral}
	\newacronym{pr}{PR}{per rectum}
	\newacronym{sc}{SC}{subcutaneous}
	\newacronym{sl}{SL}{sublingual}
	
% Glossary entries	
	\newglossaryentry{absorption}
	{
		name=absorption,
		description={the process by which medications travel through body tissues to the bloodstream}
	}

	\newglossaryentry{action}
	{
		name=action,
		description={the therapeutic effect that a medication is expected to have on the body}
	}

	\newglossaryentry{activated_charcoal}
	{
		name=activated charcoal,
		description={An oral medication that binds and adsorbs ingested toxins in the gastrointestinal (GI) tract }
	}

	\newglossaryentry{agonist}
	{
		name=agonist,
		description={medication that causes stimulation of receptors}
	}

	\newglossaryentry{antagonist}
	{
		name=antagonist,
		description={medication that binds to a receptor and blocks other medications or chemicals from attaching there}
	}

	\newglossaryentry{anaphylaxis}
	{
		name=anaphylaxis,
		description={in extreme life-threatening systemic allergic reaction that may include shock and respiratory failure}
	}

	\newglossaryentry{capsule}
	{
		name=capsule,
		description={gelatin shells filled with powdered or liquid medication}
	}

	\newglossaryentry{contraindication}
	{
		name=contraindication,
		description={when a medication would either harm the patient or have no positive effect}
	}

	% intravenous defined in ch09

	\newglossaryentry{dose}
	{
		name=dose,
		description={the amount of the medication that is given}
	}

	\newglossaryentry{enteral}
	{
		name=enteral,
		description={absorbed via the digestive system}
	}

	\newglossaryentry{generic_name}
	{
		name=generic name,
		description={the original chemical name of a medication (in contrast with one of its proprietary, or trade name); the name is not capitalized}
	}

	\newglossaryentry{hypoglycemia}
	{
		name=hypoglycemia,
		description={extremely low blood sugar}
	}

	\newglossaryentry{indication}
	{
		name=indications,
		description={reasons or conditions for which a particular medication is given}
	}

	\newglossaryentry{inhalation}
	{
		name=inhalation,
		description={administered via inhalation into the lungs}
	}

	\newglossaryentry{intramuscular}
	{
		name=\acrfull{im},
		description={administered via the muscle}
	}

	\newglossaryentry{intranasal}
	{
		name=\acrfull{in},
		description={administered into the nostril (usually via mucosal atomizer device)}
	}

	\newglossaryentry{intraosseous}
	{
		name=\acrfull{io},
		description={administered into the bone}
	}

	% intravenous defined in ch01
	
	\newglossaryentry{medication}
	{
		name=medication,
		description={substance used to treat or prevent disease or relieve pain}
	}

	\newglossaryentry{metered_dose_inhaler}
	{
		name=\acrfull{mdi},
		description={a device that delivers a consistent amount of medication using a short burst of aerosolized medicine via inhalation}
	}

	\newglossaryentry{myocardial_infarction}
	{
		name=\acrfull{mi},
		description={heart attack}
	}

	\newglossaryentry{parenteral}
	{
		name=parenteral,
		description={absorbed via means other than the digestive system}
	}

	\newglossaryentry{per_oral}
	{
		name=\acrfull{po},
		description={administered by the mouth}
	}

	\newglossaryentry{per_rectum}
	{
		name=\acrfull{pr},
		description={administered by the rectum}
	}

	\newglossaryentry{pharmacodynamics}
	{
		name=pharmacodynamics,
		description={the process by which medication works on the body}
	}

	\newglossaryentry{pharmacology}
	{
		name=pharmacology,
		description={the science of drugs, including their ingredients, preparation, uses, and actions on the body}
	}

	\newglossaryentry{proprietary_name}
	{
		name=proprietary name,
		description={the brand name that a manufacturer gives a medication; the name is capitalized}
	}

	\newglossaryentry{side_effect}
	{
		name=side effect,
		description={any action of a medication other than the desired ones}
	}

	\newglossaryentry{solution}
	{
		name=solution,
		description={liquid mixture of one or more substances that cannot be separated simply}
	}

	\newglossaryentry{sublingual}
	{
		name=\acrfull{sl},
		description={under the tongue; a medication route}
	}

	\newglossaryentry{subcutaneous}
	{
		name=\acrfull{sc},
		description={administered under the skin}
	}

	\newglossaryentry{suspension}
	{
		name=suspension,
		description={substance that does not dissolve well in liquids}
	}

	\newglossaryentry{sympathomimetic}
	{
		name=sympathomimetic,
		description={simulating sympathetic nervous action in physiological effect}
	}

	\newglossaryentry{systemic_effect}
	{
		name=systemic effect,
		description={whole-body}
	}

	\newglossaryentry{tablet}
	{
		name=tablet,
		description={contain other materials that are mixed with the medication and compressed}
	}

	\newglossaryentry{trade_name}
	{
		name=trade name,
		description={the brand name that a manufacturer gives a medication; the name is capitalized. \newline Also called proprietary name}
	}

	\newglossaryentry{transdermal}
	{
		name=transdermal,
		description={administered via the skin (alt. transcutaneous)}
	}

	\newglossaryentry{transcutaneous}
	{
		name=transcutaneous,
		description={administered via the skin (alt. transdermal)}
	}

	\newglossaryentry{unintended_effect}
	{
		name=unintended effect,
		description={effects that are undesirable but pose little risk to the patient}
	}

	\newglossaryentry{untoward_effect}
	{
		name=untoward effect,
		description={effects that can be harmful to the patient}
	}
	
	
\chapter{Principles of Pharmacology}

\subsection*{Abbreviations}
\begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries ABCD}]
	\item [\acrshort{im}] 		\acrlong{im}
	\item [\acrshort{in}] 		\acrlong{in}
	\item [\acrshort{io}] 		\acrlong{io}
	\item [\acrshort{iv}] 		\acrlong{iv}
	\item [\acrshort{mdi}] 		\acrlong{mdi}
	\item [\acrshort{mi}] 		\acrlong{mi}
	\item [\acrshort{otc}] 		\acrlong{otc}
	\item [\acrshort{po}] 		\acrlong{po}
	\item [\acrshort{pr}] 		\acrlong{pr}
	\item [\acrshort{sc}] 		\acrlong{sc}
	\item [\acrshort{sl}] 		\acrlong{sl}
\end{description}

\subsection*{Definitions}
\begin{description}	
	\item [\gls{absorption}]				\glsdesc{absorption}
	\item [\gls{action}]					\glsdesc{action}
	\item [\gls{agonist}]					\glsdesc{agonist}
	\item [\gls{antagonist}]				\glsdesc{antagonist}
	\item [\gls{capsule}]					\glsdesc{capsule}
	\item [\gls{contraindication}]			\glsdesc{contraindication}
	\item [\gls{diaphoretic}]				\glsdesc{diaphoretic}
	\item [\gls{dose}]						\glsdesc{dose}
	\item [\gls{enteral}]					\glsdesc{enteral}
	\item [\gls{generic_name}]				\glsdesc{generic_name}
	\item [\gls{hypoglycemia}]				\glsdesc{hypoglycemia}
	\item [\gls{indication}]				\glsdesc{indication}
	\item [\gls{inhalation}]				\glsdesc{inhalation}
	\item [\gls{intramuscular}]				\glsdesc{intramuscular}
	\item [\gls{intranasal}]				\glsdesc{intranasal}
	\item [\gls{intraosseous}]				\glsdesc{intraosseous}
	\item [\gls{intravenous}]				\glsdesc{intravenous}
	\item [\gls{medication}]				\glsdesc{medication}
	\item [\gls{metered_dose_inhaler}] 		\glsdesc{metered_dose_inhaler}
	\item [\gls{myocardial_infarction}] 	\glsdesc{myocardial_infarction}
	\item [\gls{parenteral}] 				\glsdesc{parenteral}
	\item [\gls{per_oral}] 					\glsdesc{per_oral}
	\item [\gls{per_rectum}] 				\glsdesc{per_rectum}
	\item [\gls{pharmacodynamics}] 			\glsdesc{pharmacodynamics}
	\item [\gls{pharmacology}] 				\glsdesc{pharmacology}
	\item [\gls{proprietary_name}] 			\glsdesc{proprietary_name}
	\item [\gls{side_effect}] 				\glsdesc{side_effect}
	\item [\gls{solution}] 					\glsdesc{solution}
	\item [\gls{sublingual}] 				\glsdesc{sublingual}
	\item [\gls{subcutaneous}] 				\glsdesc{subcutaneous}
	\item [\gls{suspension}] 				\glsdesc{suspension}
	\item [\gls{sympathomimetic}]			\glsdesc{sympathomimetic}
	\item [\gls{systemic_effect}] 			\glsdesc{systemic_effect}
	\item [\gls{tablet}] 					\glsdesc{tablet}
	\item [\gls{trade_name}] 				\glsdesc{trade_name}
	\item [\gls{transdermal}] 				\glsdesc{transdermal}
	\item [\gls{transcutaneous}] 			\glsdesc{transcutaneous}
	\item [\gls{unintended_effect}] 		\glsdesc{unintended_effect}
	\item [\gls{untoward_effect}] 			\glsdesc{untoward_effect}
\end{description}\hfill \\
\clearpage

\section{Medication Routes of Administration}
\paragraph{As an \acrshort{emt}, you will}
\begin{enumerate}
	\item Administer medications.
	\item Help patients self-administer medications.
\end{enumerate}

\subsection{Routes of Administration}
\begin{outline}	
	\1[] \textbf{absorption:} via tissues to the blood stream
	\2[] \textbf{enteral medications} enter the body \underline{through} the digestive system.
	\2[] \textbf{parenteral medications} enter the body through means \underline{other than} the digestive system.
\end{outline}		

\begin{table}[ht]
	\centering
	\caption{Routes of Medication Administration}
	
	\bgroup
	\def\arraystretch{1.25}%
	\begin{tabular}{|p{3cm}|p{1.5cm}|p{5cm}|p{4cm}|}
		\hline
		\textbf{name} 			& \textbf{abbrev.} 	& \textbf{entry point}			& \textbf{rate of absorption} \\ \hline \hline
		
		\multicolumn{4}{l}{\bfseries enteral} \\ \hline
		[per] oral				& PO				& by mouth						& slow \\ \hline 
		[per] rectal			& PR				& by rectum						& rapid \\ \hline
		sublingual 				& SL 				& under the tongue				& rapid \\ \hline
		
		\multicolumn{4}{l}{\bfseries parenteral} \\ \hline
		inhalation	 			& 					& inhaled into the lungs		& rapid \\ \hline
		intramuscular 			& IM				& into the muscle 				& moderate \\ \hline
		intranasal				& IN  				& into the nostril \newline	(via mucosal atomizer device)	
																					& rapid \\ \hline
		intraosseous 			& IO 				& into the bone					& immediate \\ \hline
		intravenous 			& IV 				& into the vein 				& immediate \\ \hline
		subcutaneous 			& SC 				& beneath the skin				& slow \\ \hline
		transcutaneous 
		\newline
		(transdermal)			&   				& through the skin  			& slow \\ \hline
	\end{tabular}
	\egroup
\end{table}\hfill \\

\pagebreak
\subsection{Medication Form}
Medication form is chosen by the manufacturer to ensure \sout{maximum profits} the following:
\begin{enumerate}
	\item Proper route of administration
	\item Timing of the medication’s release into the bloodstream
	\item Effects on the target organs or body systems
\end{enumerate}

\paragraph{Tablets, Capsules}
\begin{outline}	
	\1[] \textbf{Capsules} are gelatin shells filled with powdered or liquid medication.
	\1[] \textbf{Tablets} are contain other materials that are mixed with the medication and compressed.
\end{outline}	

\paragraph{Solutions, Suspensions}
\begin{outline}
	\1[] \textbf{Solutions} are liquid mixtures of one or more substances that \underline{cannot be separated simply}.
		\2[] Does not need to be shaken.  Can be given as an IV, IM, or SC injection
		\2[] \textbf{Example}: 
			\3[] epinephrine using an auto-injected (i.e. an Epi-Pen) \newline
	\1[] \textbf{Suspensions} substances that do not dissolve well in liquids; \underline{will separate if undisturbed/filtered}.
		\2[] Very important to shake before using!
		\2[] Injectable suspensions only via IM or SC
	\2[] \textbf{Examples}:
		\3[] activated charcoal (PO)
		\3[] some hormone shots and vaccinations (IM or SC)
		\3[] calamine lotion (topical)
\end{outline}

\paragraph{Metered-dose inhaler (MDI)}
\begin{outline}
	\1[] \textbf{Metered-dose inhalers (MDI)} direct aerosolizable liquids and fine powders through the mouth and into the lungs via inhalation.
		\2[] Delivers the \underline{same consistent dosage every time}	
		\2[] Very important to shake before using!
		\2[] \textbf{Example}:
			\3[] asthma inhalers
\end{outline}


\paragraph{Topical medications}
\begin{outline}
	\1[] Applied to \underline{skin surface}
	\1[] affects \underline{only that area}
	\1[] includes lotions, creams, and ointments
	\1[] \textbf{Examples}:
		\2[] Calamine lotion (lotion) 
		\2[] hydrocortisone cream (cream) 
		\2[] Neosporin ointment (ointment) 
\end{outline}

\paragraph{Transcutaneous medications}
\begin{outline}	
	\1[] \textbf{transcutaneous/transdermal medications} are absorbed through the skin.
	\2[] May have systematic effects (compare with \textbf{topical medications} whose effects are limited to applied area.)
	\2[] Touching will absorb medication same as patient!
	\2[] \textbf{Examples}:
		\3[] nitroglycerin paste
		\3[] adhesive patch
\end{outline}

\paragraph{Gels}
\begin{outline}	
	\1[] Semi-liquid
	\1[] Administered in capsules or plastic tubes
	\1[] \textbf{Example}:
		\2[] oral glucose
\end{outline}

\paragraph{Gases for Inhalation}
\begin{outline}	
	\1[] Outside of OR, most commonly used is \underline{oxygen}
	\1[] Usually delivered through a \underline{nonrebreathing mask} or \underline{nasal cannula}
	\1[] \textbf{Example}:
		\2[] oxygen
\end{outline}\hfill \\

\section{Administering Medication}
\subsection{The 6 "Rights" of Medication Administration}
\begin{description}
	\item [Right patient]: Patient who \underline{needs} medication = patient who \underline{receives} medication.
	\item [Right medication]: Verify that it is the correct medication and prescription.
	\item [Right dose]: Verify the \underline{form} and \underline{dose} of the medication.
	\item [Right route]: Verify the \underline{route} of the medication.
	\item [Right time]: Check the \underline{expiration date} and \underline{condition} of the medication.
	\item [Right documentation]: Document your actions and the patient’s response.
\end{description}

\subsection{Unit may carry:}
\begin{itemize}
	\item Oxygen
	\item Oral glucose
	\item Activated charcoal
	\item Aspirin
	\item Epinephrine
\end{itemize}

\subsection{Circumstances in which medications may be administered:}
\begin{outline}[enumerate]
	\1 Peer-assisted administration
	\1 Patient-assisted administration
	\1 \acrshort{emt}-administered medications
\end{outline}

\paragraph{Determined by} state and local protocols, medical control \newline
The state, department, and medical director will define which\underline{ medications are carried} on your ambulance. \hfill \\

\begin{table}[ht]
	\centering
	\caption{Advantages \& Disadvantages of Medication Administration Routes}
	
	\bgroup
	\def\arraystretch{2}%
	\begin{tabular}{p{4.5cm}|P{3cm}|p{5.5cm}}
		\hline
		\textbf{Advantages}									& 	\textbf{Route of \newline Administration}	&	\textbf{Disadvantages} \\ \hline \hline
		ease of access \newline
		comfort level 										&   PO											& 	digestive tract can be easily affected by foods, stress, and illness \newline
		speed of movement of food through the tract dramatically changes the speed of absorption \\
		
		easy to advise patients \newline
		quick absorption  									&   SL											& 	Constant evaluation of the airway \newline
		Possible choking \newline
		Not for uncooperative or unconscious patients \\
		
		quick, easy access without using vein \newline
		stable blood flow to muscle  						&   IM											&	Use of a needle (and subsequent pain) \newline
		Patients may fear pain or injury \\ \hline
	\end{tabular}
	\egroup
\end{table}
	
\afterpage{%
	\clearpage
\begin{table}[ht]
	\centering
	\caption{Drugs that can be Administered by \acrshort{emt}s}
	\bgroup
	\def\arraystretch{1.25}%
	\begin{tabular}{|p{1.5cm}|P{2cm}|P{2.5cm}|p{3cm}|p{5cm}|}
		\hline 
		\textbf{Drug}			& 	\textbf{Routes of \newline 
											Admin.}			&	\textbf{Forms}			&	\textbf{Uses} 							&	\textbf{Contraindications}	\\ \hline \hline
		
		activated\newline
		charcoal				&	PO						&	suspension				&	Reduces the amount of medication 
																							being absorbed							&	Do not give to patients with altered level of consciousness. \\ \hline
		
		oral\newline
		glucose					&	PO						&	gel, \newline
																tablet					&	Treats hypoglycemia						&	Do not give to an unconscious patients, 
		or one who cannot protect the airway. \\ \hline
		
		aspirin					&	PO						&	tablet					&	Useful during heart attack				&	Hypersensitivity to aspirin 
		Liver damage, bleeding disorder, asthma \newline
		Should not be given to children  \\ \hline
		
		nitro-\newline
		gylcerin				&	SL, \newline	
									inhalation \newline		
									(1 spray =\newline 
									1 tablet) 				&	SL tablet, \newline
																metered-dose spray		&	Relieves angina pain \newline
																							Increases blood flow \newline
																							Relaxes veins 							&	Possibility of MI, if no relief \newline
																																	Should \underline{not} be used with erectile dysfunction medications \\ \hline
		
		epine-\newline
		phrine 					&	IM						&	auto-injector			& 	Treats life-threatening anaphylaxis		&	Do not give to patients with hypertension, hypothermia, MI, or wheezing.	\\ \hline
		
		Naloxone				&	IN						& 	atomizer				&	Reverses the effects of opioid overdose	&	The effects of naloxone may not last as long as those of opioids; repeat doses may be necessary.  Can cause severe withdrawal symptoms; patients may become violent \\ \hline
		
		oxygen					&	inhalation				&	gas:
																nonrebreathing mask (preferred) \newline
																nasal cannula 			&	When a patient is not breathing, 
																							having trouble getting air				&	Ensure no open flames in vicinity \\ \hline
	\end{tabular}
	\egroup
\end{table} %
}

\afterpage{%
\clearpage
\subsection{Potential Test Questions}
\begin{outline}{enumerate}
	\1 What are enteral medication routes?  
		\1[] Per oral, per rectal, sublingual
	\1  What are parenteral medication routes?  
		\1[] intramuscular, intranasal, intraosseous, intravenous, subcutaneous, transcutaneous (transdermal)
	\1  What are the differences between capsules and tablets?  
		\1[] Capsules are gelatin shells filled with powder or liquid. Tablets having their ingredients compressed under high pressure; may contain other materials mixed with the medication.
	\1  What are the differences between solutions and suspensions?
		\1[] A solution contains substances that \textit{cannot be separated by standing or filtering}, whereas a solution will separate if undisturbed or filtered.
	\1  What should you do before administering a suspension?   
		\1[] Shake or swirl
			\2[] \textbf{Why?}
			\2[] To ensure that the patient gets receives the right amount of medication.
				\3[] \textbf{Why?}  
				\3[] Suspensions contain substances that do not dissolve well; they will separate if they stand or are filtered.
	\1  What is an example of a solution? 
		\1[] epinephrine administered via auto-injector (i.e. an EpiPen)
	\1  What is an example of a suspension? 
		\1[] Activated charcoal
	\1  What is the difference between a metered-dose inhaler and a nebulizer? 
		\1[] Nebulizer has electric components and must be recharged
	\1 What is the difference between transdermal medications and topical medications?
		\1[] topical medications affect only the intended site, transdermal medications can have systemic effects.
\end{outline}
}

\end{document}